---
title: "Counting people"
author: "Martin Summer"
format: html
editor: source
---

## Counting people

This is a project about counting people in Kenya. Like in all countries
in the world, national statistical offices describe people living in a country
by age, sex, race, education, and a number of other factors. Are these descriptions important?

In particular, does the description of a country by age groups help us better understand a
country? In this project we study the shape of Keynia based on the counts of people living in the country
by age. As in real life national statistics we are goint to produce data visualizations that 
display the data by way of barplots and other summaries. These descriptions 
begin to unpack stories about people living in this
country. Behind these numbers and graphs are real people, their sisters and brothers and parents and nieces and
nephews. The stories revealed in these data help us better understand a particular country.

::: {.callout-note}
At this stage students have not yet learned a lot in terms of technique, especially also not loading, wrangling 
and cleaning data. We thus provide them with a data table, they can still manage manually.
:::

```{r}
library(JWL)
dat <- dem_dat_by_sex_and_age_hist 

dat_kenya <- dat[dat$Country == "Kenya" &
                  dat$Year == 2022 &
                  dat$Age %in% c("0-4", "5-9", "10-14", "15-19", "20-24", "24-29", "30-34", "34-39","40-44",
                  "45-49","50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80-84", "85-89", "90-94",
                  "95-99", "100 - 105"), c("Country", "ISO2", "Year", "Sex", "Age", "POP")]

dat_kenya$Age <- factor(dat_kenya$Age, levels = c("0-4", "5-9", "10-14", "15-19", "20-24", 
                                                  "24-29", "30-34", "34-39","40-44",
                  "45-49","50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80-84", "85-89", "90-94",
                  "95-99", "100 - 105"), ordered = T)

table_kenya <- dat_kenya[ , c("Age", "Sex", "POP")]

library(tidyverse)

tab <- table_kenya  %>% pivot_wider(
names_from = Sex,
values_from= POP)

knitr::kable(tab, row.names = F)

```
```{r}
tab$"F"
```

```{r}
tab$"M"
```

```{r}
length(tab$"F")
length(tab$"M")
length(tab$Age)
```
```{r}
tab$Age
```

