# How sure can we be about what is going on: Estimates and intervals

In statistics we often make estimates based on samples of data from 
a population as we discussed in unit 3 or we make predictions
of future values of a variable using models as discussed in unit 4.
But how precise are these estimates? Do they vary a lot or only a little?
Estimates are not worth very much, if we are not able to quantify 
such uncertainty. In this unit we will learn a first technique based on
computation to do so, the bootstrap.

::: {.callout-note appearance="simple"}

## Overview

In this unit you will learn how to quantify **uncertainty intervals**
for your statistical estimates as well as for your predictions by using
computational methods. Being able
to do this is a very important part of communicating statistics.

**Bootstrapping** is the technique we will learn in this unit.
Bootstrapping a sample consists of creating new datasets of
the same size by resampling the original data with replacement.

The distribution of some inportant **sample statistics** such as the mean or the
median of a datset when computed using bootstrapping resamples are
approximated by a normal curve for larger datasets, regardless of the
shape of the original distribution.

Uncertainty intervals based on bootstrapping depend on modern computing power.
They do not require assumptions about the mathematical description of the
population distribution and do not require complex probability theory beyond the
intuitive idea of **random sampling with replacement**.

Since the computational method of the bootstrap depends on repeated computations
on random draws from a sample, you need to understand how to implement and
use **iteration** as a computational technique. In this unit you will learn
to work with **for-loops** in R and make use of them to quantify uncertainty.
Since the objects of a resample can be complex you learn also about a new
datastructure of R the R **list-class** which is very flexible in storing
data or all types and shapes.

:::

::: {.callout-important appearance="simple"}

## Keywords

**Population**: A population in statistics represents every possible
individual element we would like to measure.

**Sample**; A sample is a part or subset of the population.

Predictions with statistical models or estimates of statistics from
samples have a **margin of error**, which we need to be able to
quantify to assess the quality of a prediction or an estimate.

**Bootstrapping** is a computational technique of creating from a
given dataset new datasets of
the same size by resampling at random the original data with replacement
making the assumption that the sample is approximately distributed similarly as 
the population.

**Central Limit Theorem**: When a statistics, such as for example the mean, is
repeatedly computed from random resamples these computed values tend to be
distributed symmetrically and approximated by a normal curve for large sample 
sizes. This holds irrespective of the shape of the initial distribution
of the data.

**Lists**: Lists are an important class of objects in R. They are able to 
contain 
elements of different types like − numbers, strings, vectors and another 
list inside it.  List is created using the `list()` function.

**for-loops**: For loops in R implement iterations of computations. They
are an indispensable tool to iterate repeated computations. A for look consists
of three key elements. The preallocated output, the specification of a sequence
and the body of the for loop.

:::

How many people will there be on the planet in 2100? Nobody knows, but we
can make use of our experience from the past and our knowledge of demographic
mechanisms to build models, which support us in making systematic
predictions. You did work out such a prediction model in your last project
work of "people-count" using a regression model.

In 2020 the World Health Organisation WHO published one of it's global population
forecast. The WHO needs high quality predictions of population numbers to plan
their programs. They also use a prediction model but it is more sophisticated
than the linear model we used in the last unit. The WHO model takes many
factors into account. Here is a picture that illustrates their prediction:

```{r}
#| include: false
#| 
library(JWL)
library(tidyverse)
```

```{r}
whopredplot <- worldpop_pred %>%
  filter(country == 'WORLD') %>%
  ggplot(aes(x = year)) +
  theme_classic() +
  geom_point(aes(y = population*1e-06)) +
  geom_ribbon(aes(ymin = population_lower95*1e-06,ymax = population_upper95*1e-06),alpha = .1) + 
  labs(title = "WHO predicted world population with 95% prediction interval",
       x = "Year",
       y = "World population (Billions)")
whopredplot
```

What you see in the picture are dots indicating the WHO prediction about how many
people will be there on the planet in future years, starting from 2020, the year
this prediction was produced. The points are shown for five year intervals on the
horizontal axis and
give the population count estimate on the vertical axis. So for 
instance for 2040 the prediction is a number slightly more than 9 billion. 
The precise number is 9.198847 billion. 

But there is also a gray area surrounding the points. This area represents
a **margin of error** that belongs to this prediction. There is an upper bound and
a lower bound to this margin. The upper margin of error is 9.396987 billion and
the lower margin of error is 8.996324. So there could be about 200 million 
people more or about 400 million people less. These are huge margins. This 
is a margin of error comprising almost 2-times the size of the population (roughly 
333 Million as of 2022) of the 
United States of America. If you go further out 
into the future,
say to the year 2080, this margin of error increases. Worse than that: While
for 2080 the point estimate would still predict some sort of stabilization, the
upper margin would indicate an increase, while the lower margin would indicate
a decrease in the world population. So within the margin of error the 
predictions would even be qualitatively different.

Knowing the margin of error is important in order to get a more precise idea 
about the quality of the prediction. Anyone can make an estimate. But the real
challenge is to make an realistic assessment of the possible error. This is a
crucial element of statistics as a science and in this unit we are going to learn the
basic skills needed for making such assessments now.

The importance of having a reasonable idea about the margin of error
doe not only refer to predictions but to any other statistical 
estimate depending on limited information.

The data on population counts collected by the WHO will be mostly be
based on surveys. Let us assume the surveys are all well designed and
accurately executed. Assume that the people and institutions 
who have conducted the
surveys have been very careful to avoid biases. Then the summary statistics
calculated from the sample, such as the mean or median age of 
the people counted or any other statistics calculated from the sample, 
should be close to the
value of the statistics in the study population.

In a well designed study we expect the sample statistics, like the mean the median
the standard deviation and the interquartile range to be close to the 
mean, median, standard deviation and interquartile range of the 
study population. 

In the communication of data often only the 
summary statistics is reported and
sometimes this may be enough. But knowing the uncertainty attached to 
the estimate will be important. It can even qualitatively change the
information, as in our example, where the WHO forecast for the global population
in 2080 could mean an increase as well as a decrease of the numbers. Knowing
the exact differences between the sample and the population, or knowing the
uncertainty in a prediction may be necessary for proper statistical inference.

Let us return to a dataset we have studied before, when we discussed how we get
from information about a sample to information on the study population. Let
us take our data `socr_height_weight` about height and weight of 25000 18 year 
old people we had discussed in unit 3. We retrieve the data
from the `JWL` package and change the units of height from inches to
cm and of weight from pounds to kg. Then we store the data in an R object
which we call `dat` for simplicity.

```{r}
#| code-fold: false
library(JWL)

data <- socr_height_weight
data$Height <- data$Height*2.54
data$Weight <- data$Weight*0.4535924
```
To illustrate how the accuracy of statistics depends on sample size, let 
us take the data on human height and assume the 25000 data we have are
the population and use our function `emp_distr_height()` to draw random samples
from this population. In unit three we had drawn samples of 10, 100 and 2500.
```{r}
emp_distr_height <- function(n){
  
  data[sample(x = 1:nrow(data), size = n, replace = T), ]

  }
```
The distributions of these various samples are shown in the following figure.
```{r}
par(mfrow = c(2, 2))

h10 <- emp_distr_height(10)
hist(h10$Height, breaks = 30,
     main = "Histogram of human height N = 10",
     xlab = "Height in cm",
     ylab = "Frequency")

h100 <- emp_distr_height(100)
hist(h100$Height, breaks = 30,
     main = "Histogram of human height N = 100",
     xlab = "Height in cm",
     ylab = "Frequency")

h2500 <- emp_distr_height(2500)
hist(h2500$Height, breaks = 30,
     main = "Histogram of human height N = 2500",
     xlab = "Height in cm",
     ylab = "Frequency")

hist(data$Height, breaks = 30, 
     main = "Histogram of human height Population",
     xlab = "Height in cm",
     ylab = "Frequency")

```

As the sample size gets smaller the sample gets more and more "bumpy" 
because the histogram becomes sensitive to individual data points. Only as
the sample gets larger we come closer to the actual population distribution. This
is also intuitively clear. If somebody would tell you he had measured
10 adults from Kenya and concludes that their average height is some number, this
is less credible than in a case where the statement was based on a much larger
number of observations.

Let's go back to our problem at hand. What can we say about the mean 
height in our ficticious population based on our different samples?

```{r}
samp <- data.frame(Size = c(10, 100, 2500, 25000), 
                   Mean_Height =
                     round(
                       c(mean(h10$Height), 
                         mean(h100$Height), 
                         mean(h2500$Height), 
                         mean(data$Height)), 2 ))

knitr::kable(samp)
```

You can see from the table that with substantive larger sample sizes our 
estimation of the mean becomes more and more precise.

## Quantifying uncertainty by bootstraping: The basic idea.

In order to quantify the accuracy of our statistics we now have to work out
a thought experiment. If we assume that we repeatedly drew samples
of 25000 individuals of 18 years an each time measured their mean height,
by how much would the statistics - the mean height - vary?

If we knew how this statistics would vary, then we could say something
about the accuracy of the statistics. But the precise variablity of the
estimates depends on knowledge of the population. But this knowledge is
exactly the knowledge we do not have.

There are in principle two ways out of this circular reasoning. We could
bring in probability and mathematics to make assumptions about the
population distribution and work out the variability we would expect under these
assumptions and thus how far away our statistics in our sample 
is from the value we would theoretically expect. This is an approach you learn
in every statistics course and we will learn it too later in this course.

Here we take a second approach based on computation. We start from the
plausible assumption that the population and the sample should roughly look
the same. Since we can  not draw repeated samples from the unknown population
, we instead draw repeated samples from our sample.

Let us illustrate this idea with our height data. Let us take the sample
of 100 and draw take three bootstrap resamples from this sample.
```{r}
set.seed(123)

par(mfrow = c(2, 2))

hist(h100$Height, breaks = 30, 
     main = "Original sample N = 100",
     xlab = "Height in cm",
     ylab = "Frequency")

data <- h100

b1 <- emp_distr_height(100)

hist(b1$Height, breaks = 30,
     main = "Bootstrap 1: N = 100",
     xlab = "Height in cm",
     ylab = "Frequency")

b2 <- emp_distr_height(100)

hist(b2$Height, breaks = 30,
     main = "Bootstrap 2: N = 100",
     xlab = "Height in cm",
     ylab = "Frequency")

b3 <- emp_distr_height(100)

hist(b3$Height, breaks = 30,
     main = "Bootstrap 3: N = 100",
     xlab = "Height in cm",
     ylab = "Frequency")

```

At the top left corner of this figure you see our original sample of 100
height data from our original population of 25.000, which has a mean of
`r round(c(mean(b1$Height),2))`. Then we draw 100 data points in sequence, each time
replacing the point we have taken, and get the second distribution in the
top right corner in this way. This distribution has a mean 
of `r round(mean(b2$Height), 2)`. The distribution can only contain data points taking on the same values as in the original sample, but it will contain
different number of each value. Therefore the shape of the distribution will
vary slightly and give a slightly different mean. In our case these means
are `r round(c(mean(b1$Height), mean(b2$Height), mean(b3$Height)), 2)`.

Thus the resampling procedure gives us an idea how our statistics varies through
this procedure of resampling with replacement. Why do we need
replacement? We want to make sure that at every resample, every
data point has the same chance of ending up in the resample. This would
not be the case if we drew observations away once the have ended up in
one sample. This is known as **bootstraping**
the data. There is something "magic" about this approach, reflected in the name:
As if pulling oneself up by one's own bootstraps, here we learn about the
variability of an estimate without making assumptions about the
shape of the population distribution. We think of the sample 
as  beeing a representation
of the underlying population. If we draw random resamples we always need
to draw from the same sample of the population so that every individual 
in the sample has the same chance of ending up in the resample.

Repeating the resampling many times, say 1000 times, we get 1000 possible
estimates of the mean. Let us do this repeated resampling for the examples
of sample sizes of 10, 100, 2500 and in the end 25000, the size 
of our fictitious population.

```{r}
# Bootstrap: sample with replacement
set.seed(45356) # So you can reproduce my results

data <- socr_height_weight
data$Height <- data$Height*2.54
data$Weight <- data$Weight*0.4535924

B <- 1000 # Number of bootstrap resamples to take

n <- c(10,100,2500, 25000)

out <- vector("list", length(n))

for(j in seq_along(n)){
  
bootstrapmeans <- vector("double", length = B)

for (b in 1:B) {
  # Draw a bootstrap sample
  bootstrapsample <- sample(data$Height,n[j],replace = TRUE)
  # Compute the bootstrap estimate
  bootstrapmeans[b] <- mean(bootstrapsample)
}

out[[j]] <- bootstrapmeans

}
```

```{r}
par(mfrow = c(2, 2))

hist(out[[1]], xlim = c(168,178), ylim = c(0,400),
     main = "1000 resamples from N = 10",
     xlab = "Mean height in cm")
hist(out[[2]], xlim = c(168,178), ylim = c(0,400),
     main = "1000 resamples from  N = 100",
     xlab = "Mean height in cm")
hist(out[[3]], xlim = c(168,178), ylim = c(0,400),
     main = "1000 resamples from N = 2500",
     xlab = "Mean height in cm")
hist(out[[4]], xlim = c(168,178), ylim = c(0,400),
     main = "1000 resamples from N = 25000",
     xlab = "Mean height in cm")
```

Note that the distribution of the estimates of the mean based on the resampled
data is symmetric around the mean. The deeper mathematical reason for this
fact is the so called **Central Limit Theorem**, which says that the distribution
of sample means tends towards a normal distribution with increasing sample size
almost regardless of the shape of the original distribution of the data. We will
discuss this result in further detail later in the course.

The bootstrap distributions allow us to quantify the uncertainty connected to
the estimate of the mean based on the sampled data. For example, we can find
the range of means that contain 95 % of all the means of the bootstrap
resample and call this the 95 % uncertainty interval for the original estimates,
or the margins of error.


```{r}
mean_height <- round(c(mean(out[[1]]), mean(out[[2]]), mean(out[[3]]), mean(out[[4]])),2)
lower_margin <-round(c(
  quantile(out[[1]], probs=.025),
  quantile(out[[2]], probs=.025),
  quantile(out[[3]], probs=.025),
  quantile(out[[4]], probs=.025)),2
)

upper_margin <- round(c(
  quantile(out[[1]], probs=.975),
  quantile(out[[2]], probs=.975),
  quantile(out[[3]], probs=.975),
  quantile(out[[4]], probs=.975)), 2 
)

tab <- data.frame(mean_height, lower_margin, upper_margin)
knitr::kable(tab)
```


Also observe that the bootstrap distributions get narrower as the sample size
increases. This means that uncertainty shrinks and the estimate becomes
more precise.

One of the appeals of the bootstrap approach to quantify uncertainty is that
we were able to do this without any mathematics but the idea of drawing
observations at random. All we need is an estimate and a sampling distribution.
Thus the quantification of uncertainty is almost entirely based on
computation.

We did, so far, not show you the details of how the computations are done. We
will introduce these concepts now. They will enlarge the frontier of your R
knowledge by the concept of R's list class as well as the concept of
computational iteration.

## Doing bootstraps using R

It is now time to learn some more R, in order to do the computations
needed for a bootstrap. One convenient data structure in this context
is R's list class. The other very useful concept, which you will very
often need in computation and coding is the concept of iteration. Let
us begin with the R-list class.

### The R list class

Let us for a moment go back to the children nutrition data from the
DHS survey we discussed in unit 2 and load this dataset again and store 
them in a variable with a simpler name.

```{r}
#| code-fold: false

library(JWL)
dhs_data <- children_nutrition_data
```
In unit 2 we discussed one approach how we could compute the number of
cases of stunting across the four regions in the dataset. In the exercises
we discussed how to do the same computation more efficiently using the
`tapply()` function.

Let us introduce here a related function, called `split()`. There are usually
several ways in R how we can filter data. In unit 2 we used the `which()` function
to compute indices for the appropriate rows we wanted to select from our
dataframe. Another way could be, like this:
```{r}
#| code-fold: false

r1 <- dhs_data[dhs_data$region == "region 1", c("wt", "nt_ch_stunt")]

head(r1)

```
We can do this region by region and compute the shares also for regions 2, 3 and
4.

Here is a faster and cleaner way to achieve the same thing using `split()`:
```{r}
#| code-fold: false

regions <- split(
  dhs_data[ ,c("wt", "nt_ch_stunt", "region")], dhs_data$region
  )
lapply(regions, head)
```
Let's unpack this piece of code: You already know the syntax
`dhs_data[ ,c("wt", "nt_ch_stunt", "region")]`. This says that R should
choose from the dataframe `dhs_data` all rows and the columns with the names
"wt", "nt_ch_stunt" and "region". We could also have selected the columns by
indices.

Now this (reduced) dataframe is given as one argument 
to the function `split()`. The second argument is 
the column `dhs_data$region`. This tells the split function the criterion how
to split up the dataframe in data chunks by region, i.e. region 1, 2 etc.

The new split data are in an object called `regions` and this object has
a new class
```{r}
#| code-fold: false

class(regions)
```
This class is called in R a list. In this example the list `regions` contains
four dataframes, which we can individually access using the $-sign. We use the
`head()`function to avoid showing the entire dataframe. We applied the `heads()`
function to all elements of the list by using the function `lapply()` short for
list apply. This is a powerful functions allowing computations on elements of 
a list. Our list contains dataframes as elements. The syntax of `lapply()`takes
as input some function - what do you want to do with the list elements? - and
a list - the list containing the elements to which the function is applied. So
in this case R applies the heads function to every entry in te list. So when we

```{r}
#| code-fold: false

head(regions$"region 1")
```
Or we could alternatively use indices. But now for the `list` class we will need
double brackets, like this:
```{r}
#| code-fold: false

head(regions[[1]])
```
This retrieves the first dataframe from the list of dataframes.

Lists are particularly useful to store R objects of different types or
classes. For example, one list entry can be a dataframe, another one can consist
of characters, another of numbers and so on. The R -function `list()` creates 
a list the same way `c` creates a vector. Separate each element in the list 
with a comma: Here is a toy exmple:
```{r}
#| code-fold: false

l <- list(a = c(2,5), b = 'sky')
l
```
As we see in the toy example, we can give names to list elements. When we use
`split()` the names will be taken from the columns according to which the
data are split. If we don't like these default names we can override them by
using the `names()`function.

```{r}
#| code-fold: false

names(regions) <- c("County A", "County B", "County C", "County D")
lapply(regions, head)
```
What if we wanted the third row of the dataframe of region 1?
```{r}
#| code-fold: false

head(regions[[1]][3, ])
```
Some of you may have noticed that the $ signs are both used to refer to
columns in a dataframe as well as to elements of a list. This is no
coindcidence, because in R dataframes are lists. Each column is one element
of the list.
```{r}
#| code-fold: false

head(dhs_data[[1]])
```

### Iteration

In the project "people count" you are working on in this course, we are often
looking at countries comparatively. We looked for instance at Kenya, the 
United States and Japan and sometimes also at another country of your choice.

In coding it is an important principle to reduce duplication of code
as much as possible. This has three main benefits it focuses on the
essential parts of a code, it is easier and shorter to make changes
and it also helps reducing bugs in your code.

One tool to reduce duplication of code you have already encountered. Here, 
you learn a second tool: Iteration. When you have to do tasks 1000 and more times,
like in resampling from the same distribution, such tools are even necessary. 
These are tasks simply impossible to do by hand or by copy and paste.

#### For Loops

Imagine we have a simple dataframe, where we have four columns, each 
containing 10 random numbers from a normal distribution:
```{r}
#| code-fold: false

df <- data.frame(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```
Now, assume we want to compute the mean of every column. In this little
toy example, you could do this by hand and by copy and paste, if you
chose and solve the task as follows:
```{r}
#| code-fold: false

mean(df$a)
mean(df$b)
mean(df$c)
mean(df$d)

```
A better way to solve such tasks is to use iteration and implement
a **for-loop**. This looks like this:
```{r}
#| code-fold: false

output <- vector("double", ncol(df))  # 1. output

for (i in seq_along(df)) {            # 2. sequence
  
  output[[i]] <- mean(df[[i]])        # 3. body

  }

output
```
::: {.callout-important}

## The three components of a for loop

Every for-loop has three components:

1. The **output**: `output <- vector("double", ncol(df))`. Before you start
a loop you need to allocate space for the results computed during the loop.
This is essential for efficiency. Always preallocate space for the output
This can in general best be done using the `vector()` function of R. It has
two arguments, the class of the vector, numeric, logical, character etc. and 
the length of the vector.
2. The **sequence**: `i in seq_along(df)`. This describes what you are
looping over. Each iteration of the for loop will assign `i` to a
different value from `seq_along(df)`. 
3. The **body**:  `output[[i]] <- mean(df[[i]])`. This is the code that
does the actual work in the for loop. It i run repeatedly with each
`i`. When `i`is 1 the computation will be `output[[1]] <- mean(df[[1]])`,
when it is 2, we will have `output[[2]] <- mean(df[[2]])` etc.

:::

Let us apply this technique to a problem you had worked on before in your
project work "people count". There you were asked to find a way to
compute the mean age of a country for a situation where you do not
have individual data but data grouped by age groups. The main idae there
was to take the mid point of the age group as an approximation for the
age and multiply with the population sized in the age group. Then sum
over all age groups and divide by the total population size. Now if you do this
for many countries this is a typical situation where you want both functions
and iteration. Let us take this case as an example to show how interation
works in R.

First we load the data using the `JWL` library and assign them to an object with
a simpler name to make our life easier.

```{r}
#| code-fold: false

library(JWL)
dat <- population_statistics_by_age_and_sex
```

Now assume we want to compute the mean age of Kenya, the US, Japan and Austria, 
which happens to be my country. Instead of computing this case by case by
recycling an multiplying code, we do this using iteration.

Let's first write a function, which computes the mean for our grouped data.
The function should take the entire demographic dataframe as an input and the
country and year we choose as an argument. It can then be applied to any country
and any year by just specifying the correct argument and providing the dataframe once.

Let's start with the function:

```{r}
#| code-fold: false

mean_age <- function(country, year, data){
  
  # select the data for a given dataframe for a given country and a given year
  
  pop_dat <- data[data$Country == country & data$Year == year, ]
  
  # compute the mean age per age group by assuming that the age in a group 
  #is approximated by
  # the mid point. For example, if the age group is 0-4 we assume the mean 
  # age is 2, if it is 
  # 5 - 9 it is 7 etc. We do this by creating an appropriate sequence.
  
  mid <- seq(from = 2, to = 100, by = 5)
  
  #compute the mid age. Note that we use the R recycling rules. Mid has only 
  #half the length of 
  # pop_dat$POP, because pop_dat$POP contains first all rows for females and 
  #then for males. Once
  # mid is exhausted at age group 95-99 for females it just begins from new for 
  #men. By this we can
  # make the total sum without explicitly defining objects for each sex separately.
  
  # Compute the mean age rounded to two digits.
  
  round(sum(pop_dat$POP*mid)/sum(pop_dat$POP), 2)
  
  
}
```

Now we start with the iteration:

```{r}
#| code-fold: false

# Define the list of countries we are interested in first:

list_of_countries <- c("Kenya", "United States of America", "Japan", "Austria")

# Preallocate space for four countries: The mean is usually some decimal 
# number, so the output of our iteration will be of type double. we will compute 
# one mean age per country, so the length of the output pre-allocation needs to 
# match the length of our country list.

output <- vector("double", length(list_of_countries))  # 1. output

for( i in seq_along(list_of_countries)){               # 2: sequence
  
   # 3. Body: Note the countries change over the iteration. Data and year stay 
   #          the same.
  
  output[[i]] <- mean_age(list_of_countries[i], year = 2022, data = dat) 
 
}

```
And now, let's check:
```{r}
#| code-fold: false

output
```
Let me explain how this works. In the first step R reserves a vector for the 
entry of four doubles. Then the sequence is specified: It says that it starts at
entry 1 in the `object list_of_numbers`. This is in our example 
`list_of_numers[1]`, which must be Kenya. Then it calls the function with this
country specified as the country argument and the arguments for `year` and 
`data` set to our default values `2022` and `dat`. This output is written into
the first entry of `output` and then the next iteration starts until the end
of the country list, overwriting one component after the other with outputs
from the computation.

There are lots of variations of iteration scenarios. We won't go beyond this
most basic case here


## Standard error of a proportion

In unit 1 we computed proportions of stunted children in four of 
different regions and got these proportions as a mean to 
a series of `TRUE` or `FALSE` answers to an anthropometric
measurements and a comparison with thresholds in a WHO
incidence table. Let's retrieve these data once more:

```{r}
#| code-fold: false

library(JWL)

dhs_data <- children_nutrition_data
```
Let's compute the proportions of stunted children per region using the `tapply`function and taking
account that each observation is weighted by the `wt`, the weight, variable.
```{r}
#| code-fold: false

# create the weighted variable and attach to dataframe as an addiional column

dhs_data$weighted_nt_ch_stunt  <- dhs_data$nt_ch_stunt*dhs_data$wt

# compute total number of stunted children per region

stunted <- tapply(dhs_data$weighted_nt_ch_stunt, dhs_data$region, sum)

# compute the total number of children per region

total <- tapply(dhs_data$wt, dhs_data$region, sum)

# compute the proportions rounded to two digits

prop <- round(stunted/total,2)

prop

```

Now we want to use the bootstrap to get an idea of how precise the estimates of
these proportions are. According to the basic idea of the bootstrap we 
repeatedly resample from the empirical distribution. The estimated standard 
error of this means is then the standard deviations of these means, which are
in our case the proportions per region.

We write the bootstrap function first. Within the function we built in a
for loop to do the repeated resampling.

```{r}
#| code-fold: false

bootstrap <- function(B, data){
  
  boot <- vector("list", B)                                  # output allocation
  
  for (b in seq_along(1:B)){                                          # sequence
    
    aux <- data[sample(x = 1:nrow(data), size = nrow(data), replace = T), ]
    aux$weighted_nt_ch_stunt  <- aux$nt_ch_stunt*aux$wt
    stunted <- tapply(aux$weighted_nt_ch_stunt, aux$region, sum)
    total <- tapply(aux$wt, aux$region, sum)
    boot[[b]] <- (stunted/total)                                          # body
  
      }
 res <- do.call(rbind, boot)
 
 as.data.frame(res)
}
```

Before we look at the numerical results, let us unpack this code. Here
we have put the iteration inside the function. The function arguments
are the data and the number of bootstrap resamplings, which we call here
$B$. 

The function does $B$ random samples from the data. We use a list as an
output vector, because at every iteration we will get a named vector of
results, with four means for every region.

Now we define the sequence. We do $B$ bootstrap samples in total. You see that
the running index can have any legitimate symbol. Before we called it $i$ and 
here we call it $b$. 

Then we sample rows from our dataframe at random with replacement. This is
exactly similar to our discussion in unit 3 where we took random samples from
the dataframe of human heights. We save this in an r-object which we chose to
call `aux`here (for auxiliary). Then we do the same computation of the mean
we did before and store the result in list element $b$.

In the end we have boot filled with $B$ vectors with components showing
the means of region 1 to 4.

In the end we use an R-function to transform the list back into a dataframe.
This function is called `do.call()` and works as follows: The first argument
is a function. It specifies what `do.call()`does to every element in the list.
The function we chose is called `rbind()` (for row bind) it binds together 
rows in a dataframe. The result is stored in an R object called res and
res is  then the output. With `as.data.frame()`we tell R to give this
result the class of a dataframe. Thus our output is a dataframe
with 1000 rows and 4 columns, each column specifiying a region.

Now we can pin down the bootstrapped mean and standard deviation. Let's
give it a try. We first take 1000 bootstrap resamples from our data and
compute the weighted means or proportions for all four regions:

```{r}
#| code-fold: false

boot <- bootstrap(1000, dhs_data)

```
Now we have a dataframe with 4 columns and we want to compute the mean and sd for every column. Here we
can practice iteration once more:

```{r}
#| code-fold: false

all_means <- vector("double", 4)
all_sds <- vector("double", 4)

for(i in seq_along(boot)){
  
  all_means[[i]] <- round(mean(boot[,i]),2)
  all_sds[[i]] <- round(sd(boot[ ,i]),2)
  
  out <- data.frame(Regions = c("Region 1", "Region 2", "Regions 3", "Regions 4"), 
                    Means = all_means, SDs = all_sds)
  
  out
  
}

knitr::kable(out)
```
This should look very familiar to you by now. We allocate the space for the 
output for the means and the standard deviations. Then in the for-loop
we compute the mean for every row and the standard deviation for every
row. Finally we write everything into a dataframe. The final line
uses the `kable()` function from the `knitr` package. This allows
a particular layout for the dataframe. It gives you by the way an
alternative way to call functions from a package without running the
`library()` function first. When the package is installed functions
from the package can be called by the syntax `package_name::function_name`.

Let's look at the numbers first. We observe a few things. First, observe
that the standard deviation is largest for region 3, where the mean can vary
by four percentage points as compared to 2 percentage points in all the
other regions. While the proportion seems highest in region 2, within the
precision of the estimate it will be difficult to distinguish it from region 4.

Let us look at the histogram of the bootstrap-distribution of the means:

```{r}
par(mfrow = c(2,2))

hist_data1 <- hist(boot[ ,1], main = "Bootstrap 1", xlab = "proportion of stunted children in region 1")
#define x and y values to use for normal curve
x_values <- seq(min(boot[ ,1]), max(boot[ ,1]), length = length(boot[ ,1]))
y_values <- dnorm(x_values, mean = mean(boot[ ,1]), sd = sd(boot[ ,1])) 
y_values <- y_values * diff(hist_data1$mids[1:2]) * length(boot[ ,1]) 

#overlay normal curve on histogram
lines(x_values, y_values, lwd = 2, col = "red")

hist_data2 <- hist(boot[ ,2], main = "Bootstrap 2", xlab = "proportion of stunted children in region 2")
#define x and y values to use for normal curve
x_values <- seq(min(boot[ ,2]), max(boot[ ,2]), length = length(boot[ ,2]))
y_values <- dnorm(x_values, mean = mean(boot[ ,2]), sd = sd(boot[ ,2])) 
y_values <- y_values * diff(hist_data2$mids[1:2]) * length(boot[ ,2]) 

#overlay normal curve on histogram
lines(x_values, y_values, lwd = 2, col = "red")

hist_data3 <- hist(boot[ ,3], main = "Bootstrap 3", xlab = "proportion of stunted children in region 3")
#define x and y values to use for normal curve
x_values <- seq(min(boot[ ,3]), max(boot[ ,3]), length = length(boot[ ,3]))
y_values <- dnorm(x_values, mean = mean(boot[ ,3]), sd = sd(boot[ ,3])) 
y_values <- y_values * diff(hist_data3$mids[1:2]) * length(boot[ ,3]) 

#overlay normal curve on histogram
lines(x_values, y_values, lwd = 2, col = "red")

hist_data4 <- hist(boot[ ,4], main = "Bootstrap 4", xlab = "proportion of stunted children in region 4")
#define x and y values to use for normal curve
x_values <- seq(min(boot[ ,4]), max(boot[ ,4]), length = length(boot[ ,4]))
y_values <- dnorm(x_values, mean = mean(boot[ ,4]), sd = sd(boot[ ,4])) 
y_values <- y_values * diff(hist_data4$mids[1:2]) * length(boot[ ,4]) 

#overlay normal curve on histogram
lines(x_values, y_values, lwd = 2, col = "red")
```
We have overlayed all four distributions with a normal curve. You see that the
fit is pretty good. This is no accident, as we will learn the deeper 
reason for this in the coming units.

## Regression and the Pearson height data

One of the appeals of the bootstrap is that it is very general. Any estimate can
be bootstrapped. All we need is an estimate and a sampling
distribution. The bootstrap can be performed algorithmically without
the need of a formal probability model. We can apply this idea to any
situation where we need to quantify uncertainty.

In unit 5 we fitted a regression line to the heights data of fathers and son 
collected by Pearson in 1903. Let us come back to this analysis now.

```{r}
#| code-fold: false

library(JWL)
dat <- pearson
mod <- lm(Son ~ Father, data = dat)
mod
```
Our estimated coefficient were $33.89$ for the intercept and  $0.514$ for the 
slope. How confident can we be about the position
of the regression line? We could answer this question by bootstrapping.

Let us take a bootstrap resample of the 1078 Father and Son pairs with replacement.
From the data we fit a least squares line. Repeating this many times we can
compute a 95% interval for the coefficient. Let's use R to implement this idea.

```{r}
#| code-fold: false

B <- 1000
output <- vector("list", length = B)

for(i in seq_along(1:B)){
  
  aux <- dat[sample(x = 1:nrow(dat), size = nrow(dat), replace = T), ]
  mod <- lm(Son ~ Father, data = aux)
  
  output[[i]] <- mod$coefficients
}

res <- do.call(rbind, output)
 
result <- as.data.frame(res)
```

This code should be no mystery for you by now. One point is worth mentioning
though. When we call the `lm()` function to do a least suqare fit of
the regression line, we store the output in an R object called mod. When R
performs the computations implemented in `lm()` it does not only compute
the coefficients of the regression line but a lot of more different
things and stores them in a list. The value of the coefficients are 
stored in a list element with the name `coeffcients`. Thus we store the
value of the coefficients by assigning `mod$coefficients` to the object where
we store the results.

When we compute the mean and the standard deviation of the coefficient, we get:

```{r}
#| code-fold: false

round(mean(result$Father),2)
```

95 % of the values of the coefficient are between the 2.5 and the 97.5 
percentile. This can be computed using the `quantile()` function:

```{r}

#| code-fold: false

round(quantile(result$Father, c(0.025, 0.975) ),2)
```

This gives you a sense of the range where the coefficient value of this 
can fall with high confidence.

## Uncertainty quantification beyond the bootstrap

Bootstrapping provides an intuitive way of assessing uncertainty of estimates.
The method is computer intensive and needs computing technology and capacity.
Since these tools are becoming available more widely and at ever lower cost,
this is an attractive approach when we need to judge the uncertainty we
should attach to our estimates.

The bootstrap is a very powerful method but always keep the following
things in mind:

::: {.callout-important}

## Things to keep in mind about the bootstrap

1. Try to start with a sufficiently large random sample. Note that the
sample and hence the resample will only resemble the population distribution
if the sample is large enough.

2. To approximate the distribution of a statistics, like a mean, a median etc.
replicate the resampling procedure very often. To get a decent approximation
a few thousand resamples will be good. As a rule of thumb you could use 5000
or 10000, dependning in the computational ressoruces available to you.

3. The bootstrap percentile method works well for estimating the population
mean of median based on a large random sample. It is not expected to work very well when the minimum or maximum should be estimated or when we look for very
low or very high percentiles or other parameters that are strongly influenced
by extreme observations. It also does not work well when the distribution of the
statistics is not symmetric or well approximated by a normal curve. It also does
not work well when the sample is very small say lower than 10 or 15.


:::

Bootstrapping reaches its limit when very large quantities of data are
involved. Here we have reached the limits of what we can learn about
statistics without taking resort to the (mathematical) theory of
probability.

In such cases, when we need to deal with very large quantities of data,
bootstrapping is too clumsy. In these cases we need to resort to the elegant 
and effective mathematics of uncertainty which is based on the theory of
probability, which we explore in the next unit. We will only
introduce the basic ideas without using too much mathematics. These
ideas will be necessary to understand how we can leverage
probability theory to quantify uncertainty.



## Exercises

### Exercises:

:::{.callout-note icon=false}
## Exercise 1: Thirsty students

Assume you are interested in the amount of water the student 
coming to the JWL study center consume in a day. In particular 
you want to learn about the proportion of students that drink 
more than 2 liters of water per day. You take a random sample 
of 100 students in the students directory of the study center 
and obtain from them the information about how much water they 
drink in a day at the study center. You store your 
results in a dataframe you call  `water`. The dataframe 
has one column called `amount` which stores the number of liters each of the students drink.

1. Write one line of R-code to compute the proportion of 
students who drank more than 2 liters of water. Hint: You 
can write this line of code also when you are not at the 
computer. You can solve this exercise by thinking about 
what you would write, if you had an R console or a Jupyter book opened in front of you.

2. Write one line of R-code to compute a single bootstrap resample from the water dataframe.

:::

:::{.callout-note icon=false}

## Exercise 2: Check your understanding of why the bootstrap works

1. When we conduct a bootstrap resample, what size resample 
should we draw from our sample? Why? Hint: There is no unique 
answer to the size question only rules of thumb. When you 
think about the why, go back to consult our discussion on why 
random sampling works in unit 2.

2. Why do we need to sample with replacement?

3. When we conduct a bootstrap resample, what is the 
underlying assumption/reasoning for resampling from
our sample? Why does it work?


:::

### Exercises R

:::{.callout-note icon=false}
## Exercise 1: A toy problem for practicing iteration

Load the pearson height data from the `JWL`package 
and transform the data to standard units using a 
for loop.
Hint: You will need a function and an interation 
over the columns of the dataframe.

:::

:::{.callout-note icon=false}
## Exercise 2: Bootstrapped interquartile range of height data

To practice the bootstrapping technique we learned in this unit, let 
us load the human height data `socr_height_weight` once again 
from the JWL package. Take a sample of 5000 of the height data 
and compute the interquartile range. Implement a bootstrap 
with 10000 resamples to quantify the uncertainty in the sample 
inter quartile range for the inter quartile range, i.e. the 
difference between the 25 and the 75 percentile of the height 
distribution. Compute the range in which 95 % of the 
inter-quartile-ranges lie. You can leave the data for this 
exercise in the original units of inches. Hint: For the 
interquartile range consult the passage on percentiles 
in unit 3. If you do 10000 bootstraps this might take a 
while in times of computation. As long as the `In [*]` cell 
has a star in the brackets and not a number, the computation 
is still going on. So you will need a bit of patience for 
computing this exercise.

:::


## Project People count: The present and future of humanity in pictures and numbers

### How precise are our estimates of the mean age ?

In our project in unit 2 we estimated the mean age of the
population in Kenya, the United States and Japan as well as in
your own country in the year 2022. In this project we will leverage our
knowledge of the bootstrap to attach some uncertainty interval
to our estimate. Let us use Kenya as an example in this part
of the project for. You may also choose your own country if
you are more interested in this particular case.

1. Filtering the data for Kenya (or any other country): This is a repetition
of what we have done 
already previously. Take the data `population_statistics_by_age_and_sex` 
from the `JWL` package and store it in an R object. Check whether the data 
you get in this way is what you want.

2. Implement a bootstrap for the mean age in Kenya (or the country 
of your choice) using your previous
results and the techniques you learned in this unit and report the bounds that
contain 95 % of these estimates based on 5000 resamples from the 
appropriate distribution.

3. What kind of data would you need to get a more precise estimate?
What do you think? Explain in a few sentences.
