# Modelling relationship: Prediction using regression

We ended the last chapter with Pearson's historical data on the
heights of fathers and sons to introduce the concept of a scatter-plot.

```{r}

library(JWL)
dat <- pearson

plot(dat, 
     pch = 19, 
     cex = 0.3, 
     xlab = "Father's height (inches)", 
     ylab = "Son's height (inches)",
     main = "Pearson's height data from 1903")
abline(0,1, lty = 2)
```

We discussed how we could use the positive association in this scatter-plot to 
predict the height of a son from the height of the father. One intuition might be
to use the straight dashed line that run along the diagonal. If we would use this
line as a model for prediction, each adult son would be predicted to show the same 
height as his father. But we can do better than that and improve on this choice.

For any straight line we might choose, each data-point will give rise to a **residual** which
is the size of the error were we to use the line as prediction.

```{r}
plot(dat, 
     pch = 19, 
     cex = 0.3, 
     xlab = "Father's height (inches)", 
     ylab = "Son's height (inches)",
     main = "Pearson's height data from 1903")
abline(0,1, lty = 2)
abline(reg = lm(dat$Son ~ dat$Father), col = "red", lwd = 2)
points(63.1, 74.3, pch = 19, col = "red")
lines(c(63.1, 63.1), c(66.3264, 74.3), col = "red", lty = 2)
```

Suppose for instance that we took the father with a height of 63.1 inches and predicted the height
of the son according to the model described by the red line, we would have predicted the son to
have height 66.3 inches. In the data the son of the father with height 63.1 inches is in fact 74.3
inches. This is marked out as the red point in the plot. The length of the dashed line from the
red point to the red line is the residual.

The idea of prediction using models of straight lines is to make these residuals small. We
will learn a method how to do this. This approach to prediction is called regression and we
will learn how it works in this chapter.