# What causes what?

<!-- ## Contents -->

<!-- In this lecture we discuss what causation means in a statistical sense, why we need be careful to distinguish between causation and correlation. -->

<!-- 1. Causation in a statistical sense means that when we intervene, the chances of different outcomes vary systematically. -->
<!-- 2. Causation is difficult to establish statistically, but well designed randomized trials are the best available framework. -->
<!-- 3.Principles that helped clinical trials to identify effects. -->
<!-- 4, Observational data may have background factors influencing the apparent relationships between exposure and outcome which may be either observed confounders or lurking factors. -->
<!-- 5. Statistical methods do not suspend judgment which is always required for the confidence with which causation can be claimed. -->

<!-- ## Outcome -->

<!-- This is a conceptually difficult topic. It is, however, not technically difficult but it will need lots of examples. Fortunately there are many good (and bad) real world examples that I hope will stick with the students as reference examples after the course. Students should understand the idea of randomized trials and observational studies as well as the general ideas of comparison in statistical analysis. -->

The last two centuries brought enormous advances in the expansion of education around the
world. Quantitative measures such as literacy rates, school enrollment, the share of 
education spending in gdp all point into the same direction: While access to education
had been the privilege of a very small elite throughout history, things have changed
in the direction of a very broad participation. 

This is a result of both economic
necessities of modern production as well as of political struggles for human rights and
equality. Today education is perceived as a right and it is seen as a government duty to
provide broad access to basic education. A certain amount of schooling is even compulsory in
most countries, although not all the countries have the capacity to broadly enforce such obligations.

Despite all of these developments and efforts in the 
expansion of education there are still about 8 % of children worldwide 
which are not in school.
Also progress has been heterogeneous across the globe, with literacy rates
among the youth of sub Saharan Africa - for example - still below 50 %.^[The data and an interesting analysis can be found in @Roser2021 at https://ourworldindata.org/children-not-in-school]

Education policies have thus been widely debated 
in international policy circles and 
have spurred debates as well as to 
seemingly unresolvable controversies about 
what is needed to further improve education for 
those who are still behind and precisely which policies would and
would not work towards this goal.

For example, the majority view in international policy circles has been that the focus of education policy should be to
get children into the class room and provide the
necessary teachers. This entails policies to 
make sure that children have
access to a school nearby as well as the 
free provision of basic education. Once this is
assured, according to this approach, the rest will
more or less follow automatically.

But enrollment does not automatically mean that all the children enrolled actually attend.
For example, we know from the DHS, the 
demographic an health survey you learned about
in previous lectures, that there is a gap 
between enrollment and attendance in the countries 
that have the most urgent need for
improved education policies. From the
surveys we also know that the absence is not 
necessarily driven by the fact that the
children are needed at home. 

Look at this graph compiled from DHS data provided by Our World in Data:

```{r}
library(JWL)
plot(enrol_attend_dat$Enrolment, enrol_attend_dat$Attendance, 
     main = ("Enrollment versus Attendance"), xlab = ("Enrollment rate primary school"), ylab = ("Attendance rate primary school"), pch = 16)

abline(1,1, col = "blue")
```


Another aspect that was debated is that enrollment does not necessarily mean that children
actually learn something or that they learn at a sufficient quality. This problem of poor 
learning outcomes in many parts of the globe was demonstrated in many cases.

Does this mean these policies are wrong, as their critics claim. They point out that there
is no point in all this educational supply if  there is no demand 
for it, taking lack of attendance and lack of sufficient learning 
outcomes as evidence. If there were demand, these
critics argue, supply would follow and the demand arises by itself as soon as there are
high enough returns to education.

Why can't controversies like these not simply be resolved by looking at the data?

## Correlation does not imply causation

Many data sources we have just rely on education policy outcomes somebody has observed and
recorder, so called observational studies. For instance there is a large literature, 
mainly from before the mid 1990ies in education research, which tried to answer 
the following question: Do increased financial resources for public 
schools improve educational outcomes as measured by standard education achievement tests?

A look across countries would - for instance - give the following picture.

```{r}
plot(expenditure_outcome_dat$Expenditure, 
     expenditure_outcome_dat$Outcome, 
     main = ("Average learning outcomes by total education expenditure per capita"), 
     xlab = ("Public and private per capita expenditure on education (PPP, constant 2011-intl $)"), 
     ylab = ("Average harmonized learning outcome in 2005 - 2015"), pch = 16)

axis(1, at = seq(500, 3500, by = 500))
```


Indeed a study by @Hedgesetal1994 confirmed patterns like these. There is strong
evidence that education spending and educational outcomes are positively correlated.
Remember our discussion of measuring the relation between two variables in section 3.5?

But does this evidence support the conclusion that there is a causal relation between 
increased educational spending and student outcomes? There is a straightforward reason why
this can not be the case. We compare very different students or schools accross countries in
the scatterplot we produced before. In the study by @Hedgesetal1994 the variables that were
related were students from different households across different schools. For
the general point we want to make here, this is immaterial. What matters is that simply comparing outcomes of students across countries or students across schools with different
levels of spending does not tell us whether the different spending levels are causal for the
variation in outcomes we observe. There may be many other differences between countries and
students that are shaping this relationship.

Statisticians have summarized this observation in the mantra 
that **"correlation does not imply causation"**. If we want to 
establish a causal relation from data we need to work harder.

## What does it mean that a variable is "causal"?